---
title: "presence_absence_heatmaps_METABOLIC"
author: "Taruna Schuelke"
date: "08 July 2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) # this means your code will be generated alongside your output (i.e., data visuals/model summaries)
knitr::opts_chunk$set(message = FALSE) # this means the message from commands won't be printed out in the .html file
knitr::opts_chunk$set(warning = FALSE) # this means the warning from commands won't be printed out in the .html file
```

## Step 1. Load libraries
```{r libraries}
# Load required libraries
library(ggplot2)
library(reshape2)
library(dplyr)
library(readxl)
library(GGally)
library(tidyr)
library(pheatmap)
library(tidyverse)
```


### Step 2. Read in the functions presence/absence data. Make sure the "Present" & "Absent" strings are replaced with 1 and 0, respectively.

```{r CSVfiles}
metabolic_function <- read.csv("METABOLIC_result_worksheet2.tsv",  sep = "\t")
simple_IDs <- read.csv("Bin_ID_to_Mag_ID.csv")
```

### Step 3. Always `View` your data
```{r}
View(metabolic_function)
```


### Step 4. Check if the *Function* column contains unique functions
```{r}
# Extract the Function column
function_column <- metabolic_function$Function

# Check for uniqueness
unique_functions <- unique(function_column)

# Compare the length of unique values with the original column length
is_unique <- length(unique_functions) == length(function_column)

# Print the result
if (is_unique) {
  print("The Function column contains all unique strings.")
} else {
  print("The Function column contains duplicate strings.")
}

# Optionally, display the duplicate values
if (!is_unique) {
  duplicate_values <- function_column[duplicated(function_column)]
  print("Duplicate values:")
  print(unique(duplicate_values))
}
```

### Step 5. If there are duplicate functions, combine the functions with the gene abbreviations and create a new column with this info
```{r}
# Combine "Function" and "Gene.abbreviation" columns to create "Function_GeneAbbreviation"
metabolic_function <- metabolic_function %>%
  mutate(Function_GeneAbbreviation = paste(Function, "[", Gene.abbreviation, "]"))
# Move "Function_GeneAbbreviation" to be the first column
metabolic_function <- metabolic_function %>%
  select(Function_GeneAbbreviation, everything())
```

### Step 6. Remove the old "Function", "Category", and "Gene.abbreviation" columns
```{r}
metabolic_function <- metabolic_function %>%
  select(-Function, -Category, -Gene.abbreviation)
```


### Step 7. Clean-up the Mag_IDs which are the columns
```{r}
colnames(metabolic_function) <- gsub("X", "", colnames(metabolic_function))
colnames(metabolic_function) <- gsub(".Function.presence", "", colnames(metabolic_function))
```

### Step 8. We need to transpose the data so that the MAGs are in rows not columns
```{r}
metabolic_function_transpose <- as.data.frame(t(metabolic_function))
#View(metabolic_function_transpose)
```

### Step 9. Assign the first row as column names and then remove it
```{r}
colnames(metabolic_function_transpose) <- metabolic_function_transpose[1, ]
metabolic_function_transpose <- metabolic_function_transpose[-1, ]
metabolic_function_transpose <- type.convert(metabolic_function_transpose, as.is = TRUE)
```

### Step 10. Make the Bin_IDs an actual column so we can match the Bin_IDs to simpler Mag_IDs
```{r}
# Convert row names to a column
metabolic_function_transpose <- cbind(RowName = rownames(metabolic_function_transpose), metabolic_function_transpose)

# Rename the new column if you want a specific name
names(metabolic_function_transpose)[1] <- "Bin_ID"

```

### Step 11. Merge the cleaned functions dataframe with the file containing simple Mag_IDs and taxonomic information
```{r}
metabolic_function_transpose_w_simple_Mag_IDs <- merge(metabolic_function_transpose, simple_IDs, by="Bin_ID" )
#View(metabolic_function_transpose_w_simple_Mag_IDs)
```

### Step 12. Sort the data frame
```{r}
# Extract the number from the Mag_ID column and create a sorting key
metabolic_function_transpose_w_simple_Mag_IDs <- metabolic_function_transpose_w_simple_Mag_IDs %>%
  mutate(SortingKey = as.numeric(gsub(".*\\.(\\d+)\\s.*", "\\1", Mag_Fam)))

# Sort the dataframe by the new sorting key column
metabolic_function_transpose_w_simple_Mag_IDs <- metabolic_function_transpose_w_simple_Mag_IDs %>%
  arrange(SortingKey)  

# Display the first few rows of the sorted and subsetted dataframe
colnames(metabolic_function_transpose_w_simple_Mag_IDs)

```

### Step 13. Subset the columns to be used to plot the heatmap
```{r}
metabolic_function_transpose_w_simple_Mag_IDs_subset <- metabolic_function_transpose_w_simple_Mag_IDs %>%
  select(107, 2:105)
```

### Step 14. Always `View` your data
```{r}
View(metabolic_function_transpose_w_simple_Mag_IDs_subset)
```

### Step 15. Extract data to create the heatmap
```{r}
# Extract the numeric data for the heatmap
significant_mags <- c("MAG.1 (f_SHLQ01)",
                   "MAG.2 (f_SHLQ01)",
                   "MAG.4 (f_Amoebophilaceae)",
                   "MAG.5 (f_Amoebophilaceae)",
                   "MAG.7 (f_Microscillaceae)",
                   "MAG.19 (f_Flavobacteriaceae)",
                   "MAG.29 (f_Schleiferiaceae)",
                   "MAG.41 (f_Promineofilaceae)",
                   "MAG.43 (f_Promineofilaceae)",
                   "MAG.45 (f_Microcoleaceae)",
                   "MAG.52 (f_Cyanobiaceae)",
                   "MAG.67 (f_Maricaulaceae)",
                   "MAG.87 (f_Rhodobacteraceae)",
                   "MAG.89 (f_Rhodobacteraceae)",
                   "MAG.103 (f_Alteromonadaceae)",
                   "MAG.106 (f_Alteromonadaceae)",
                   "MAG.107 (f_Alteromonadaceae)",
                   "MAG.108 (f_Vibrionaceae)",
                   "MAG.109 (f_Vibrionaceae)",
                   "MAG.112 (f_Cellvibrionaceae)",
                   "MAG.116 (f_DSM-6294)",
                   "MAG.117 (f_DT-91)",
                   "MAG.119 (f_Endozoicomonadaceae)",
                   "MAG.136 (f_Marinomonadaceae)")

# Filter the dataframe to include only the specified MAGs
heatmap_data <- metabolic_function_transpose_w_simple_Mag_IDs_subset %>%
  filter(Mag_Fam %in% significant_mags)

# Set rownames
rownames(heatmap_data) <- heatmap_data$Mag_Fam

# Remove the Mag_Fam column
heatmap_data <- heatmap_data %>%
  select(-Mag_Fam)
```

### Step 16. Extract data to create the heatmap
```{r}
# Generate the heatmap
svg(file = "metabolic_heatmap_sargPatch_MAGs.svg", width = 40, height = 30)
pheatmap(as.matrix(heatmap_data),
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         color = c("white", "black"),
         main = "Presence/Absence of Each MAG",
         display_numbers = FALSE,
         annotation_names_row = TRUE,
         cellwidth = 20,  # Adjust as necessary to achieve the desired size
         cellheight = 20,  # Adjust as necessary to achieve the desired size
         border_color = "gray20",  # Very light gray
         fontsize_row = 14,  # Increase row label font size
         fontsize_col = 14)  # Increase column label font size
dev.off()
```

